---
title: "Simulating data in R"
author: "Stirling Coding Club"
subtitle: stirlingcodingclub.github.io/simulating_data
date: "14 December 2022"
output:
  beamer_presentation:
    theme: "default"
    colortheme: "default"
    fonttheme: "default"
  ioslides_presentation: default
  slidy_presentation: default
header-includes:
- \usepackage{hyperref}
colorlinks: true
linkcolor: blue
urlcolor: blue
---


## Why simulate data?

Simulating data uses generating random data sets with known properties using code (or some other method). This can be useful in a lot of contexts.

\begin{itemize}
\item Better understand statistical methods
\item Plan ahead for actual data collection
\item Visualise data sets and distributions
\item Run some statistical analyses (randomisation)
\end{itemize}

## How can data be simulated in R?

Random data with different properties can be generated in R using several base R functions.

\begin{itemize}
\item `runif` generates random values from a uniform distribution. \pause
\item `rnorm` generates random values from a normal distribution. \pause
\item `rpois` generates random values from a poisson distribution. \pause
\item `rbinom` generates random values from a binomial distribution. \pause
\item `sample` samples values from any given vector with or without replacement.
\end{itemize}

Other R packages, such as the MASS library, can simulate full data sets with pre-defined correlation stuctures.

## The `runif` function in R

```{r}
rand_unifs <- runif(n = 10000, min = 0, max = 1);
```

```{r, echo = FALSE}
library(knitr);
print(rand_unifs[1:40]);
```

## The `runif` function in R

```{r}
rand_unifs <- runif(n = 10000, min = 0, max = 1);
```

```{r, echo = FALSE}
hist(rand_unifs, xlab = "Random value (X)", col = "blue",
     main = "", cex.lab = 1.5, cex.axis = 1.5);
```

## The `runif` function in R

\small

```
int verify_seed(int x){
  x=abs(x) % 30000;
  return(++x);
 } /* Easy way of getting seeds */

double as183(int seeds[]){
  double unidev; /* Code below verifies the 3 seeds */
  seeds[0] = verify_seed(seeds[0]);
  seeds[1] = verify_seed(seeds[1]);
  seeds[2] = verify_seed(seeds[2]);
  /* Code below gets a decimal to be added to unidev */
  seeds[0] = (171 * seeds[0]) % 30269;
  seeds[1] = (172 * seeds[1]) % 30307;
  seeds[2] = (170 * seeds[2]) % 30323;
  /* unidev gets a random uniform number between zero and one */
  unidev = seeds[0]/30269.0 + seeds[1]/30307.0 + seeds[2]/30323.0;
  /* Return just the decimal, subtract integer of unidev */
  return(unidev - (int)unidev);
} /* We now have one random uniform number */
```

## The `rnorm` function in R

```{r}
rand_rnorms <- rnorm(n = 10000, mean = 0, sd = 1);
```

```{r, echo = FALSE}
print(rand_rnorms[1:40]);
```

## The `rnorm` function in R

```{r}
rand_rnorms <- rnorm(n = 10000, mean = 0, sd = 1);
```

```{r, echo = FALSE}
hist(rand_rnorms, xlab = "Random value (X)", col = "blue",
     main = "", cex.lab = 1.5, cex.axis = 1.5);
```

## The `rpois` function in R

```{r}
rand_rpois <- rpois(n = 10000, lambda = 4);
```

```{r, echo = FALSE}
print(rand_rpois[1:40]);
```

## The `rpois` function in R

```{r}
rand_rpois <- rpois(n = 10000, lambda = 4);
```

```{r, echo = FALSE}
hist(rand_rpois, xlab = "Random value (X)", col = "blue",
     main = "", cex.lab = 1.5, cex.axis = 1.5);
```

## The `rbinom` function in R

```{r}
rand_rbinom <- rbinom(n = 100, size = 1, prob = 0.5);
```

```{r, echo = FALSE}
print(rand_rbinom[1:40]);
```

## The `rpois` function in R

```{r}
rand_rbinom <- rbinom(n = 100, size = 1, prob = 0.5);
```

```{r, echo = FALSE}
hist(rand_rbinom, xlab = "Random value (X)", 
     col = "blue",
     main = "", cex.lab = 1.5, 
     cex.axis = 1.5);
```


## Using `sample` in R

Create a vector of numbers from which to sample.

```{r}
my_sample_vec <- 1:10;
```

```{r, echo = FALSE}
print(my_sample_vec);
```
\pause
Use `sample` to randomly sample numbers from `my_sample_vec`

```{r}
my_sample <- sample(x = my_sample_vec, size = 4);
```

```{r, echo = FALSE}
print(my_sample);
```


## Using `sample` in R

Can sample with or without replacement.

```{r}
sample_no_replace <- sample(x = my_sample_vec, 
                            size = 10, replace = FALSE);
```

```{r, echo = FALSE}
print(sample_no_replace);
```
\pause
```{r}
sample_replace <- sample(x = my_sample_vec, 
                         size = 10, replace = TRUE);
```

```{r, echo = FALSE}
print(sample_replace);
```

## Using `sample` in R

Can also change the probabilities of being sampled

```{r}
# Vector values must sum to 1
pr_vector  <- c(0, 0, 0, 0, 0, 
                0.2, 0.2, 0.2, 
                0.2, 0.2);
new_sample <- sample(x = 1:10, size = 10, 
                     replace = TRUE, 
                     prob = pr_vector);
```

```{r, echo = FALSE}
print(new_sample);
```

## Using `sample` in R

Can also sample strings instead of numbers

```{r}
species   <- c("species_A", "species_B", "species_C");
sp_sample <- sample(x = species, size = 12, 
                    replace = TRUE, 
                    prob = c(0.5, 0.25, 0.25) 
                    );
```

```{r, echo = FALSE}
print(sp_sample);
```


## Building a simple simulated dataset

\small

```{r}
N         <- 12;
species   <- c("species_A", "species_B");
sp_sample <- sample(x = species, 
                    size = N, replace = TRUE);
sp_mass   <- rnorm(n = N, mean = 100, sd = 4);
for(i in 1:N){
  if(sp_sample[i] == "species_A"){
    sp_mass[i] <- sp_mass[i] + rnorm(n = 1, 
                                     mean = 2, sd = 1);
  }
}
sim_data  <- data.frame(sp_sample, sp_mass);
```

## Building a simple simulated dataset

```{r, echo = FALSE}
kable(sim_data);
```

## Building a simple simulated dataset

```{r}
t.test(sp_mass ~ sp_sample, data = sim_data);
```


## Building a simple simulated dataset

\small

```{r}
N         <- 120;
species   <- c("species_A", "species_B");
sp_sample <- sample(x = species, size = N, 
                    replace = TRUE);
sp_mass   <- rnorm(n = N, mean = 100, sd = 4);
for(i in 1:N){
  if(sp_sample[i] == "species_A"){
    sp_mass[i] <- sp_mass[i] + rnorm(n = 1, 
                                     mean = 2, sd = 1);
  }
}
sim_data  <- data.frame(sp_sample, sp_mass);
```

## Building a simple simulated dataset

```{r}
t.test(sp_mass ~ sp_sample, data = sim_data);
```

## Setting a seed gives the same numbers

Try it once

```{r}
set.seed(10);
rnorm(n = 10);
```

Try it again

```{r}
set.seed(10);
rnorm(n = 10);
```
